Cosmological constraints from the redshift dependence of the Alcock-Paczynski effect: significant improvement in the dark energy figure of merit

### 2017-11-03 Draft/Draft1.8.tex
minor changes

%[Xiao-Dong~Li, Cristiano G. Sabiu, Yuting Wang, Changbom~Park, Gong-bo Zhao, Hyunbae Park, Arman Shafieloo, Juhan Kim and Sungwook E. Hong]
\author{Xiao-Dong~Li$^{1,2}$, Cristiano G. Sabiu$^3$, Yuting Wang$^4$, Changbom~Park$^2$, Gong-bo Zhao$^4$, Hyunbae Park$^3$, Arman Shafieloo$^3$, Juhan Kim$^5$ and Sungwook E. Hong$^3$}
%{ Xiao-Dong Li$^{1,2,\dagger}$, Changbom Park$^{1}$, Cristiano G. Sabiu$^{2}$, Juhan Kim$^{3,1,\star}$, Sungwook E. Hong$^{1}$\\
\affil{$^1$School of Physics and Astronomy, Sun Yat-Sen University, Guangzhou 510297, P.R.China}
\affil{$^2$School of Physics, Korea Institute for Advanced Study, 85 Heogiro, Dongdaemun-gu, Seoul 02455, Korea}
\affil{$^3$Korea Astronomy and Space Science Institute, 776, Daedeokdae-ro, Yuseong-gu, Daejeon, 34055, Korea}
\affil{$^4$National Astronomy Observatories, Chinese Academy of Science, Beijing, 100012, P.R.China}
\affil{$^5$Center for Advanced Computation, Korea Institute for Advanced Study, 85 Hoegi-ro, Dongdaemun-gu, Seoul 02455, Korea}

@@

\begin{abstract}
We derive tight constraints on dynamical dark energy 
from the Alcock-Paczynski (AP) test via the redshift dependence of clustering on scales of 6 -- 40 $h^{-1}\rm Mpc$
to place constraints on the expansion history of the Universe.
We use the 1,133,326 galaxies from the Sloan Digital Sky Survey Data Release 12, 
spanning the redshift range $0.15<z<0.69$.
%Concentrating on the Chevallier-Polarski-Linder parametrization of dark energy, $w=w_0+w_a\frac{z}{1+z}$,
%while improving upon earlier works \citep{Li2016},
%we present a novel use of the Alcock-Paczynski (AP) test via clustering difference between redshift bins on scales of 6-40 $h^{-1}\rm Mpc$.
%This differential measure is shown to be a robust statistical tool that mitigates many systematic effects, 
%allowing for competitive and unbiased constrains to be placed on cosmological parameters.
We improve the methodology of Li et al. to enable efficient exploring of three or more cosmological parameters,
and apply it to the Chevallier-Polarski-Linder parametrization of dark energy, $w=w_0+w_a{z}/({1+z})$.
In combination with the CMB, BAO, SNIa and $H_0$ from Cepheid data, we obtain
$\Omega_m = 0.301 \pm 0.008,\ w_0 = -1.042 \pm 0.067,\ $ and $w_a = -0.07 \pm 0.29$ (68.3\% CL). 
Adding our new AP measurements to the aforementioned results 
reduces the error bars by $\sim$30 -- 40\%
%and reduces the area of $w_0-w_a$ contour by 50\%.
and improves the dark energy figure of merit by a factor of 2. 
We check the robustness of the results.
%The AP method may %be combined with the standard BAO method,
%greatly increase the power of ongoing and future galaxy surveys in probing dark energy.
\end{abstract}

@@

\begin{figure*}
   \centering{
   %\includegraphics[width=9cm,natwidth=4,natheight=4]{figCPL_a.pdf}
   \includegraphics[width=8.95cm]{fig_xi_cosmo1.eps}
   \includegraphics[width=8.95cm]{fig_xi_cosmo2.eps}
   }
   \caption{\label{fig_xi}
   $\hat\xi_{\Delta s}(\mu)$ measured from the SDSS BOSS DR12 galaxies in six redshift bins (three in LOWZ and three in CMASS),
   assuming the $\Omega_m=0.26$ $\Lambda$CDM cosmology and a more dark energy dominated cosmology, $\Omega_m=$0.11 $\Lambda$CDM.
   Measurements in the six redshift bins and their redshift evolution with respective to the first bin of LOWZ are plotted.
   In the $\Omega_m$=0.11 cosmology, the shapes of $\hat\xi_{\Delta s}(\mu)$ are different from the $\Omega_m=0.26$ cosmology results,
   and the difference changes with redshift;
   a large redshift evolution of $\hat\xi_{\Delta s}(\mu)$ is detected in this cosmology, 
   indicating that it is not likely to be the underlying true cosmology of our universe.
   The measurements in the HR4 mock catalogues (always in the $\Omega_m=0.26$ $\Lambda$CDM cosmology; plotted in green color)
   match the shape of curves measured from observational data, 
   indicating that the simulation well reproduces the FoG and Kaiser effects.
   For the $\Omega_m$=0.11 cosmology, we also plot the approximate 2PCFs (red dashed lines) 
   inferred using the technique described in Sec. \ref{sec:approx_2pcf}.
   The error induced in the approximation procedure is very small.
   }
\end{figure*}

@@

{\bf TBA}
As an illustration,
Figure \ref{fig_xi} shows how we us the above procedure to distinguish different cosmologies.
Here we plot the value of $\hat\xi_{\Delta s}(\mu)$, as long as its redshift evolution, 
measured from the BOSS DR12 galaxies in six redshift bins.
Two $\Lambda$CDM cosmologies are adopted, 
one with $\Omega_m=0.26$, and the other with obviously wrong value of $\Omega_m=0.11$.

The shape of $\hat \xi_{\Delta s}(\mu)$ is very different from a flat curve, 
due to the apparent anisotropy produced by the peculiar motion of galaxies.
In the $\Omega_m=0.11$ cosmology, 
the shapes of $\hat\xi_{\Delta s}(\mu)$ are different from the measurements in the $\Omega_m=0.26$ cosmology,
and the amount of difference systematically evolves with redshift.
So we observe a large redshift evolution of $\hat\xi_{\Delta s}(\mu)$, 
indicating that it is not likely to be the underlying true cosmology of our universe.

The green curves denote the 2PCFS measured from the HR4 mock catalogues
(we plot the {\it correct} measurement in the simulation cosmology, i.e. the $\Omega_m=0.26$ $\Lambda$CDM).
%The small redshift evolution of $\hat \xi_{\Delta s}(\mu)$
The simulation results can well match the results from observational data,
indicating that the FoG \citep{FOG} and Kaiser \citep{Kaiser1987} effects are both well reproduced.
Since there is no AP effect in the simulation measurements,
all detected redshift evolution should be due to effects other than the cosmological effect; 
so they are adopted as an estimation of the systematic effects of the method.
The amount of systematics reaches 4 -- 6\% in the 6th redshift bin,
and is much smaller ($\lesssim2\%$) in the other bins.

@@

The above procedure is illustrated in Figure \ref{fig_method}.
Using the relations given by Equation \ref{eq:smumap},
we obtained the distribution of number counts in cosmologies other than the fiducial cosmology.
To ensure the accuracy of mapping, the number counting is conducted using 5 times smaller pixels (the small red pixels), 
and grouped together to infer the number counts in the standard resolution (the large blue dashed pixels).

%We checked that the approximate method can reproduce the results presented in \cite{Li2016} quite well.
{\bf TBA}
Figure \ref{fig_xi} plots the approximate 2PCFS in the $\Omega_m=0.11$ $\Lambda$CDM cosmology.
We find that, the approximation procedure only introduces a $\lesssim0.5\%$ error in the $\hat\xi_{\Delta s}(\mu)$, 
which is 10 times smaller than the intrinsic noise (the possion noise and cosmic variance) in $\hat\xi_{\Delta s}(\mu)$.
%So we expect that the approximate method can reproduce the results of precise method quite well.
So it should be precise enough to use the approximate 2PCF in the statistical analysis.

@@

Figure \ref{fig_con} shows the 68.3\% and 95.4\% CL likelihood contours in the $\Omega_m - w_0$  and $w_0 - w_a$ planes,
derived from the CMB+BAO, CMB+BAO+SNIa+$H_0$ and CMB+BAO+SNIa+$H_0$+AP, respectively.
The overlapping of different contours suggest that they are consistent with each other.

@@

We can not derive effective constraint on the parameter space from a combination of CMB+BAO.
Combining the four external techniques, i.e. CMB+BAO+SNIa+$H_0$, leads to effective constraints on all parameters.
The statistical mean values and 68.3\% uncertainties of these parameters are
\begin{eqnarray}
&\Omega_m = 0.309 \pm 0.010,\\
&w_0 = -0.938 \pm 0.109,\\
&w_a = -0.38 \pm 0.41,
\end{eqnarray}
while adding our AP method to this combination further tightens the constraints and yield to
\begin{eqnarray}
&\Omega_m = 0.301 \pm 0.008,\\
&w_0 = -1.042 \pm 0.067,\\
&w_a = -0.07 \pm 0.29.
\end{eqnarray}
The error bars are dramatically reduced by 30 -- 40\%,
and the contour areas are reduced by 50\%, i.e., the dark energy figure of merit was improved by 100\%. 
Notice that the AP constraints come from the BOSS DR12 data, which is already used in the BAO analysis.
So the doubling of the figure of merit comes at no additional cost or alteration to data size,
thus greatly improves the overall cost-benefit of the large cosmological redshift surveys.

It can be also noted that, after adding the AP method, the central value of $w_a$ moved significantly towards zero.
This means the result becomes more consistent with a cosmological constant dark energy component having no evolution.
Figure \ref{fig_wz} shows the redshift evolution of $w(z)$ derived from the cosmological constraints.
Adding AP tightens the constraints and reduces the redshift evolution of $w$ (tilt of $w(z)$).

@@

The AP method presented in this work has many advantages over the 
`traditional' methods using galaxy clustering.
Since it works with the redshift evolution of the anisotropic clustering signal it significantly reduces the effect of systematics. 
It avoids the great difficulty in accurate modeling of the RSD, non-linear clustering and galaxy bias.
%It is applicable on much smaller clustering scales, allowing for the inclusion of more $k$-modes and hence more information. 
Our method can use the galaxy clustering signal down to
much smaller scales compared to other methods,
providing much higher statistics.

>>>>

%[Xiao-Dong~Li, Cristiano G. Sabiu, Yuting Wang, Changbom~Park, Gong-bo Zhao, Hyunbae Park, Arman Shafieloo, Juhan Kim and Sungwook E. Hong]
\author{Xiao-Dong~Li$^{1,2,4}$, Cristiano G. Sabiu$^3$, Yuting Wang$^4$, Changbom~Park$^2$, Gong-bo Zhao$^4$, Hyunbae Park$^3$, Arman Shafieloo$^3$, Juhan Kim$^5$ and Sungwook E. Hong$^3$}
%{ Xiao-Dong Li$^{1,2,\dagger}$, Changbom Park$^{1}$, Cristiano G. Sabiu$^{2}$, Juhan Kim$^{3,1,\star}$, Sungwook E. Hong$^{1}$\\
\affil{$^1$School of Physics and Astronomy, Sun Yat-Sen University, Guangzhou 510297, P.R.China}
\affil{$^2$School of Physics, Korea Institute for Advanced Study, 85 Heogiro, Dongdaemun-gu, Seoul 02455, Korea}
\affil{$^3$Korea Astronomy and Space Science Institute, 776, Daedeokdae-ro, Yuseong-gu, Daejeon, 34055, Korea}
\affil{$^4$National Astronomy Observatories, Chinese Academy of Science, Beijing, 100012, P.R.China}
\affil{$^5$Center for Advanced Computation, Korea Institute for Advanced Study, 85 Hoegi-ro, Dongdaemun-gu, Seoul 02455, Korea}

@@

\begin{abstract}
We derive tight constraints on dynamical dark energy 
from the Alcock-Paczynski (AP) test via the redshift dependence of galaxy clustering on scales of 6 -- 40 $h^{-1}\rm Mpc$
to place constraints on the expansion history of the Universe.
We use the 1,133,326 galaxies from the Sloan Digital Sky Survey Data Release 12, 
spanning the redshift range $0.15<z<0.69$.
%Concentrating on the Chevallier-Polarski-Linder parametrization of dark energy, $w=w_0+w_a\frac{z}{1+z}$,
%while improving upon earlier works \citep{Li2016},
%we present a novel use of the Alcock-Paczynski (AP) test via clustering difference between redshift bins on scales of 6-40 $h^{-1}\rm Mpc$.
%This differential measure is shown to be a robust statistical tool that mitigates many systematic effects, 
%allowing for competitive and unbiased constrains to be placed on cosmological parameters.
We improve the methodology of Li et al. to enable efficient exploring of three or more cosmological parameters,
and apply it to the Chevallier-Polarski-Linder parametrization of dark energy, $w=w_0+w_a{z}/({1+z})$.
In combination with the CMB, BAO, SNIa and $H_0$ from Cepheid data, we obtain
$\Omega_m = 0.301 \pm 0.008,\ w_0 = -1.042 \pm 0.067,\ $ and $w_a = -0.07 \pm 0.29$ (68.3\% CL). 
Adding our new AP measurements to the aforementioned results 
reduces the error bars by $\sim$30 -- 40\%
%and reduces the area of $w_0-w_a$ contour by 50\%.
and improves the dark energy figure of merit by a factor of 2. 
We check the robustness of the results.
%The AP method may %be combined with the standard BAO method,
%greatly increase the power of ongoing and future galaxy surveys in probing dark energy.
\end{abstract}

@@

\begin{figure*}
   \centering{
   %\includegraphics[width=9cm,natwidth=4,natheight=4]{figCPL_a.pdf}
   \includegraphics[width=8.95cm]{fig_xi_cosmo1.eps}
   \includegraphics[width=8.95cm]{fig_xi_cosmo2.eps}
   }
   \caption{\label{fig_xi}
   $\hat\xi_{\Delta s}(\mu)$ measured from the SDSS BOSS DR12 galaxies in six redshift bins (three in LOWZ and three in CMASS),
   assuming the $\Omega_m=0.26$ $\Lambda$CDM cosmology and a dynamical dark energy cosmology with $\Omega_m=0.26$, $w_0=-1$ and $w_a=-2$.
   Measurements in the six redshift bins and their redshift evolution with respective to the first bin of LOWZ are plotted.
   In the dynamical dark energy cosmology, the shapes of $\hat\xi_{\Delta s}(\mu)$ are different from the $\Omega_m=0.26$ cosmology results,
   and the difference changes with redshift;
   a larger redshift evolution of $\hat\xi_{\Delta s}(\mu)$ is detected in this cosmology, 
   indicating that it is less likely to be the underlying true cosmology of our universe.
   The measurements in the HR4 mock catalogues (always in the $\Omega_m=0.26$ $\Lambda$CDM cosmology; plotted in green color)
   match the shape of curves measured from observational data, 
   indicating that the simulation well reproduces the FoG and Kaiser effects.
   For the dynamical dark energy cosmology, we also plot the approximate 2PCFs (red dashed lines) 
   inferred using the technique described in Sec. \ref{sec:approx_2pcf}.
   The error induced in the approximation procedure is very small.
   }
\end{figure*}

@@

As an illustration,
Figure \ref{fig_xi} shows how we us the above procedure to distinguish different cosmologies.
Here we plot the value of $\hat\xi_{\Delta s}(\mu)$ (upper panels) as well as its redshift evolution (lower panels), 
measured from the BOSS DR12 galaxies in six redshift bins.
Two cosmologies are adopted, 
one is the $\Omega_m=0.26$ $\Lambda$CDM cosmology, 
and the other is a dynamical dark energy cosmology with $\Omega_m=0.26$, $w_0=-1$ and $w_a=-2$.

The shape of $\hat \xi_{\Delta s}(\mu)$ is very different from a flat curve, 
due to the apparent anisotropy produced by the peculiar motion of galaxies.
In the dynamical dark energy cosmology, 
the shapes of $\hat\xi_{\Delta s}(\mu)$ are different from the measurements in the $\Omega_m=0.26$ cosmology,
and the amount of difference systematically evolves with redshift.
We observe a larger redshift evolution of $\hat\xi_{\Delta s}(\mu)$ compared with $\Lambda$CDM cosmology, 
indicating that it is less likely to be the underlying true cosmology of our universe.

The green curves denote the 2PCFS measured from the HR4 mock catalogues
(we plot the {\it correct} measurement in the simulation cosmology, i.e. the $\Omega_m=0.26$ $\Lambda$CDM).
%The small redshift evolution of $\hat \xi_{\Delta s}(\mu)$
The simulation results can match well the results from observational data,
indicating that the FoG \citep{FOG} and Kaiser \citep{Kaiser1987} effects are both well reproduced.
Since there is no AP effect in the simulation measurements,
all detected redshift evolution should be due to effects other than the cosmological effect; 
so they are adopted as an estimation of the systematic effects of the method.
The amount of systematics reaches 4 -- 6\% in the 6th redshift bin,
and is much smaller ($\lesssim2\%$) in the other bins.

@@

The above procedure is illustrated in Figure \ref{fig_method}.
Using the relations given by Equation \ref{eq:smumap},
we obtained the distribution of number counts in cosmologies other than the fiducial cosmology.
We ensure the accuracy of the remapping by performing the pair counting using 5 times smaller pixels (the small red pixels), 
and regrouping these together to infer the number counts at the desired resolution (the large blue dashed pixels).

%We checked that the approximate method can reproduce the results presented in \cite{Li2016} quite well.
Figure \ref{fig_xi} plots the approximate 2PCFS in the dynamical dark energycosmology.
We find that, the approximation procedure only introduces a $\lesssim0.5\%$ error in the $\hat\xi_{\Delta s}(\mu)$, 
which is 10 times smaller than the intrinsic noise (the possion noise and cosmic variance) in $\hat\xi_{\Delta s}(\mu)$.
%So we expect that the approximate method can reproduce the results of precise method quite well.
So it should be precise enough to use the approximate 2PCF in the statistical analysis.

@@

Figure \ref{fig_con} shows the 68.3\% and 95.4\% CL likelihood contours in the $\Omega_m - w_0$  and $w_0 - w_a$ planes,
derived from the CMB+BAO, CMB+BAO+SNIa+$H_0$ and CMB+BAO+SNIa+$H_0$+AP, respectively.
The overlapping of the various contours suggest that they are consistent with each other.

@@

We can not derive effective constraint on the parameter space from a combination of CMB+BAO.
Combining the four external techniques, i.e. CMB+BAO+SNIa+$H_0$, leads to effective constraints on all parameters.
The statistical mean values and 68.3\% uncertainties of these parameters are
\begin{eqnarray}
&\Omega_m = 0.309 \pm 0.010,\\
&w_0 = -0.938 \pm 0.109,\\
&w_a = -0.38 \pm 0.41,
\end{eqnarray}
while adding our AP method to this combination further tightens the constraints leading to
\begin{eqnarray}
&\Omega_m = 0.301 \pm 0.008,\\
&w_0 = -1.042 \pm 0.067,\\
&w_a = -0.07 \pm 0.29.
\end{eqnarray}
The error bars are dramatically reduced by 30 -- 40\%,
and the contour areas are reduced by 50\%, i.e., the dark energy figure of merit is improved by 100\%. 
Notice that the AP constraints come from the BOSS DR12 data, which is already used in the BAO analysis.
So the doubling of the figure of merit comes at no additional cost or alteration to data size,
thus greatly improves the overall cost-benefit of the large cosmological redshift surveys.

It can be also noted that, after adding the AP method, the central value of $w_a$ moved significantly towards zero.
This means the result becomes more consistent with a cosmological constant dark energy component having no evolution.
Figure \ref{fig_wz} shows the redshift evolution of $w(z)$ derived from the cosmological constraints.
Adding the new AP results tightens the constraints and reduces the redshift evolution of $w$ (tilt of $w(z)$).

@@

The AP method presented in this work has many advantages over the 
`traditional' methods using galaxy clustering.
Since it works with the redshift evolution of the anisotropic clustering signal it significantly reduces the effect of systematics. 
It avoids the great difficulty in accurate modeling of the RSD, non-linear clustering and galaxy bias.
%It is applicable on much smaller clustering scales, allowing for the inclusion of more $k$-modes and hence more information. 
Our method can use the galaxy clustering signal down to
much smaller scales compared to other methods, thus including many more k-modes, maximising the information gain.
